---
title: "Polar plots"
output: html_notebook
---


# Load dataset 
I've made a sample dataset called "polar.csv" 
The dataset has two groups, basal and treated, each with 4 samples. Metabolites are organized in rows for now.
```{r}
library(readr)
#desktop
#mydata <- read_csv(file="C:/Users/eflam/Box Sync/Data/Human HF cohort/raw metabolomics data and pipeline inputs/LC carnitines normalized Blood.csv")
#laptop
mydata <- read_csv(file="C:/Users/Emily/Desktop/Box Sync/Data/Human HF cohort/raw metabolomics data and pipeline inputs/metabolite groups for polar plots/final combined cohort/nucleotide synthesis no EC.csv")
#summary(mydata)

```


Normalize row-wise to mean 0 and variance 1 (you may do without, but it's like when you do a heatmap, if you're values are not scaled it's gonna be hard to compare them.)
```{r}

mydata[,2:ncol(mydata)] <- apply(mydata[,2:ncol(mydata)],2,as.numeric)

mydata.t <- t(mydata[,2:ncol(mydata)])
#mydata.t <- apply(mydata.t,2,as.numeric,na.rm=TRUE)
mydata.norm <- t(scale(mydata.t))

mean(mydata.norm[1,],na.rm=TRUE) # this should be close to 0
var(mydata.norm[1,],na.rm=TRUE) # this should be 1
class(mydata.norm) # it's a matrix now, we want it back to dataframe
mydata.norm <- data.frame(mydata.norm, Metabolites= mydata$Metabolites)

class(mydata.norm)
```


Calculate mean,SE and upper and lower limit for the error bars of each group - PLASMA
```{r}
library(tidyverse)
library(plotrix) # we need this package to use the function std.error
Data <- mydata.norm %>% rowwise() %>% mutate(
                                        Mean.basal = mean(c(NF_X1381_Norm, NF_X1421_Norm, NF_X1454_Norm,	NF_X1473_Norm, NF_X1500_Norm, NF_X1528_Norm, NF_X1578_Norm, NF_X1604_Norm,	NF_X1622_Norm, NF_X1635_Norm,	NF_X1644_Norm,	NF_X1647_Norm, NF_X1648_Norm,	NF_X1650_Norm, NF_X1660_Norm,	NF_X1669_Norm,	NF_X1670_Norm, NF_X1673_Norm,	NF_X1676_Norm, NF_X1678_Norm,	NF_X1681_Norm, NF_X1690_Norm, NF_X1696_Norm,	NF_X1698_Norm, NF_X1700_Norm,	NF_X1705_Norm, NF_X1708_Norm, NF_X1711_Norm,	NF_X1713_Norm, NF_X1716_Norm,	NF_X1718_Norm, NF_X1723_Norm, NF_X1727_Norm,	NF_X1728_Norm, NF_X1731_Norm,	NF_X1732_Norm, NF_X1739_Norm, NF_X1741_Norm,	NF_X1744_Norm, NF_X1747_Norm,	NF_X1750_Norm,	NF_X1754_Norm, NF_X1755_Norm,	NF_X1758_Norm, NF_X1759_Norm,	NF_X1762_Norm,	NF_X1763_Norm, NF_X1766_Norm,	NF_X1767_Norm, NF_X1771_Norm, NF_X1779_Norm,	NF_X1785_Norm, NF_X1789_Norm), na.rm=T), 
                                        Mean.treated = mean(c(HF_X1466_Norm, HF_X1538_Norm, HF_X1618_Norm,	HF_X1657_Norm, HF_X1658_Norm,	HF_X1668_Norm, HF_X1686_Norm, HF_X1689_Norm,	HF_X1692_Norm, HF_X1704_Norm,	HF_X1706_Norm,	HF_X1709_Norm, HF_X1736_Norm, HF_X1737_Norm, HF_X1742_Norm, HF_X1764_Norm,	HF_X1768_Norm, HF_X1770_Norm,	HF_X1776_Norm, HF_X1780_Norm,	HF_X1787_Norm, HF_X1792_Norm, HF_X1793_Norm,	HF_X1794_Norm), na.rm=T),
                                        SE.basal = std.error(c(NF_X1381_Norm, NF_X1421_Norm, NF_X1454_Norm,	NF_X1473_Norm, NF_X1500_Norm, NF_X1528_Norm, NF_X1578_Norm, NF_X1604_Norm,	NF_X1622_Norm, NF_X1635_Norm,	NF_X1644_Norm,	NF_X1647_Norm, NF_X1648_Norm,	NF_X1650_Norm, NF_X1660_Norm,	NF_X1669_Norm,	NF_X1670_Norm, NF_X1673_Norm,	NF_X1676_Norm, NF_X1678_Norm,	NF_X1681_Norm, NF_X1690_Norm, NF_X1696_Norm,	NF_X1698_Norm, NF_X1700_Norm,	NF_X1705_Norm, NF_X1708_Norm, NF_X1711_Norm,	NF_X1713_Norm, NF_X1716_Norm,	NF_X1718_Norm, NF_X1723_Norm, NF_X1727_Norm,	NF_X1728_Norm, NF_X1731_Norm,	NF_X1732_Norm, NF_X1739_Norm, NF_X1741_Norm,	NF_X1744_Norm, NF_X1747_Norm,	NF_X1750_Norm,	NF_X1754_Norm, NF_X1755_Norm,	NF_X1758_Norm, NF_X1759_Norm,	NF_X1762_Norm,	NF_X1763_Norm, NF_X1766_Norm,	NF_X1767_Norm, NF_X1771_Norm, NF_X1779_Norm,	NF_X1785_Norm, NF_X1789_Norm),na.rm=T),
                                        SE.treated = std.error(c(HF_X1466_Norm, HF_X1538_Norm, HF_X1618_Norm,	HF_X1657_Norm, HF_X1658_Norm,	HF_X1668_Norm, HF_X1686_Norm, HF_X1689_Norm,	HF_X1692_Norm, HF_X1704_Norm,	HF_X1706_Norm,	HF_X1709_Norm, HF_X1736_Norm, HF_X1737_Norm, HF_X1742_Norm, HF_X1764_Norm,	HF_X1768_Norm, HF_X1770_Norm,	HF_X1776_Norm, HF_X1780_Norm,	HF_X1787_Norm, HF_X1792_Norm, HF_X1793_Norm,	HF_X1794_Norm),na.rm=T),
                                        Down.basal = Mean.basal - SE.basal,
                                        Up.basal = Mean.basal + SE.basal,
                                        Down.treated = Mean.treated - SE.treated,
                                        Up.treated = Mean.treated + SE.treated,na.rm=T)
class(Data)





```


TISSUE
```{r}
library(tidyverse)
library(plotrix) # we need this package to use the function std.error
Data <- mydata.norm %>% rowwise() %>% mutate(
                                        Mean.basal = mean(c(NF_X1052_Norm, NF_X1155_Norm, NF_X1160_Norm, NF_X1200_Norm, NF_X1221_Norm, NF_X1294_Norm, NF_X1310_Norm, NF_X1325_Norm, NF_X1356_Norm, NF_X1381_Norm, NF_X1421_Norm, NF_X1454_Norm, NF_X1473_Norm, NF_X1500_Norm, NF_X1528_Norm, NF_X1578_Norm, NF_X1580_Norm, NF_X1604_Norm,	NF_X1622_Norm, NF_X1635_Norm,	NF_X1644_Norm, NF_X1647_Norm, NF_X1648_Norm,	NF_X1650_Norm, NF_X1660_Norm,	NF_X1669_Norm, NF_X1670_Norm, NF_X1673_Norm,	NF_X1676_Norm, NF_X1681_Norm,	NF_X1690_Norm, NF_X1696_Norm, NF_X1698_Norm, NF_X1700_Norm, NF_X1705_Norm,	NF_X1708_Norm, NF_X1711_Norm, NF_X1713_Norm,	NF_X1716_Norm, NF_X1718_Norm,	NF_X1723_Norm,	NF_X1727_Norm, NF_X1728_Norm, NF_X1731_Norm, NF_X1732_Norm, NF_X1739_Norm,	NF_X1741_Norm, NF_X1744_Norm, NF_X1747_Norm, NF_X1750_Norm,	NF_X1754_Norm, NF_X1758_Norm, NF_X1759_Norm,	NF_X1763_Norm, NF_X1766_Norm,	NF_X1767_Norm,	NF_X1771_Norm, NF_X1779_Norm,	NF_X1785_Norm, NF_X1789_Norm), na.rm=T), 
                                        Mean.treated = mean(c(HF_X1007_Norm, HF_X1018_Norm, HF_X1048_Norm, HF_X1051_Norm, HF_X1104_Norm, HF_X1127_Norm, HF_X1142_Norm, HF_X1179_Norm, HF_X1275_Norm, HF_X1282_Norm, HF_X1290_Norm, HF_X1292_Norm, HF_X1298_Norm, HF_X1300_Norm, HF_X1304_Norm, HF_X1466_Norm, HF_X1538_Norm, HF_X1618_Norm, HF_X1657_Norm, HF_X1658_Norm, HF_X1668_Norm, HF_X1686_Norm, HF_X1689_Norm, HF_X1692_Norm, HF_X1704_Norm, HF_X1706_Norm, HF_X1709_Norm, HF_X1720_Norm, HF_X1725_Norm, HF_X1736_Norm, HF_X1737_Norm, HF_X1742_Norm, HF_X1764_Norm, HF_X1768_Norm, HF_X1770_Norm, HF_X1776_Norm, HF_X1787_Norm, HF_X1792_Norm, HF_X1793_Norm, HF_X1794_Norm), na.rm=T),
                                        SE.basal = std.error(c(NF_X1052_Norm, NF_X1155_Norm, NF_X1160_Norm, NF_X1200_Norm, NF_X1221_Norm, NF_X1294_Norm, NF_X1310_Norm, NF_X1325_Norm, NF_X1356_Norm, NF_X1381_Norm, NF_X1421_Norm, NF_X1454_Norm, NF_X1473_Norm, NF_X1500_Norm, NF_X1528_Norm, NF_X1578_Norm, NF_X1580_Norm, NF_X1604_Norm,	NF_X1622_Norm, NF_X1635_Norm,	NF_X1644_Norm, NF_X1647_Norm, NF_X1648_Norm,	NF_X1650_Norm, NF_X1660_Norm,	NF_X1669_Norm, NF_X1670_Norm, NF_X1673_Norm,	NF_X1676_Norm, NF_X1681_Norm,	NF_X1690_Norm, NF_X1696_Norm, NF_X1698_Norm, NF_X1700_Norm, NF_X1705_Norm,	NF_X1708_Norm, NF_X1711_Norm, NF_X1713_Norm,	NF_X1716_Norm, NF_X1718_Norm,	NF_X1723_Norm,	NF_X1727_Norm, NF_X1728_Norm, NF_X1731_Norm, NF_X1732_Norm, NF_X1739_Norm,	NF_X1741_Norm, NF_X1744_Norm, NF_X1747_Norm, NF_X1750_Norm,	NF_X1754_Norm, NF_X1758_Norm, NF_X1759_Norm,	NF_X1763_Norm, NF_X1766_Norm,	NF_X1767_Norm,	NF_X1771_Norm, NF_X1779_Norm,	NF_X1785_Norm, NF_X1789_Norm),na.rm=T),
                                        SE.treated = std.error(c(HF_X1007_Norm, HF_X1018_Norm, HF_X1048_Norm, HF_X1051_Norm, HF_X1104_Norm, HF_X1127_Norm, HF_X1142_Norm, HF_X1179_Norm, HF_X1275_Norm, HF_X1282_Norm, HF_X1290_Norm, HF_X1292_Norm, HF_X1298_Norm, HF_X1300_Norm, HF_X1304_Norm, HF_X1466_Norm, HF_X1538_Norm, HF_X1618_Norm, HF_X1657_Norm, HF_X1658_Norm, HF_X1668_Norm, HF_X1686_Norm, HF_X1689_Norm, HF_X1692_Norm, HF_X1704_Norm, HF_X1706_Norm, HF_X1709_Norm, HF_X1720_Norm, HF_X1725_Norm, HF_X1736_Norm, HF_X1737_Norm, HF_X1742_Norm, HF_X1764_Norm, HF_X1768_Norm, HF_X1770_Norm, HF_X1776_Norm, HF_X1787_Norm, HF_X1792_Norm, HF_X1793_Norm, HF_X1794_Norm),na.rm=T),
                                        Down.basal = Mean.basal - SE.basal,
                                        Up.basal = Mean.basal + SE.basal,
                                        Down.treated = Mean.treated - SE.treated,
                                        Up.treated = Mean.treated + SE.treated,na.rm=T)
class(Data)





```

Gather values in long format
Change format to long and calculate mean+ and - SE in separate columns
```{r}
data.means <- Data[,c("Metabolites", "Mean.basal", "Mean.treated")] %>%
  gather(Sample, Mean, -Metabolites)
data.SE <- Data[,c("Metabolites",  "SE.basal", "SE.treated")] %>%
  gather(Sample, SE, -Metabolites)
data.UP <- Data[,c("Metabolites", "Up.basal", "Up.treated")] %>%
  gather(Sample, Up, -Metabolites)
data.DOWN <- Data[,c("Metabolites", "Down.basal", "Down.treated")] %>%
  gather(Sample, Down, -Metabolites)

dataMeanSe <- cbind(data.means,data.SE, data.UP, data.DOWN)
dataMeanSe$Sample <- factor(dataMeanSe$Sample)
dataMeanSe$Treatment <- factor(ifelse(grepl("basal", dataMeanSe$Sample), "NF","HF"))

#drop some unneeded columns
data.polar <- dataMeanSe[c(1,3,6,9,12, 13)]
```


To generate the polar plot, first we work with a normal line plot and add error bars as ribbon (or any style you want)
```{r}

a <- data.polar %>% 
  ggplot(aes(x=Metabolites, y = Mean, group=Treatment)) +
  #geom_boxplot(aes(group = Group, colour = Genotype, fill = Treatment, alpha = 0.2))
  geom_ribbon(aes(ymin = Down, ymax = Up, fill = Treatment), alpha = 0.15) +
  scale_fill_manual(values= c("blue", "red")) +
  geom_line(aes(colour = Treatment)) +
  scale_color_manual(values= c("blue", "red")) +
  theme_light() +
  theme(panel.grid.minor = element_blank())  

plot(a)
```

Now we change coordinate system
```{r}
a + coord_polar() 
```

