---
title: "Polar plots"
output: html_notebook
---


# Load dataset 
I've made a sample dataset called "polar.csv" 
The dataset has two groups, basal and treated, each with 4 samples. Metabolites are organized in rows for now.
```{r}
library(readr)
mydata <- read_csv(file.choose())
summary(mydata)
```


Normalize row-wise to mean 0 and variance 1 (you may do without, but it's like when you do a heatmap, if you're values are not scaled it's gonna be hard to compare them.)
```{r}

mydata.t <- t(mydata[,2:ncol(mydata)])
mydata.norm <- t(scale(mydata.t))

mean(mydata.norm[1,]) # this should be close to 0
var(mydata.norm[1,]) # this should be 1
class(mydata.norm) # it's a matrix now, we want it back to dataframe
mydata.norm <- data.frame(mydata.norm, Metabolites= mydata$Metabolites)

class(mydata.norm)
```


Calculate mean,SE and upper and lower limit for the error bars of each group
```{r}
library(tidyverse)
library(plotrix) # we need this package to use the function std.error
Data <- mydata.norm %>% rowwise() %>% mutate(
                                        Mean.basal = mean(c(basal_1, basal_2, basal_3, basal_4), na.rm=T), 
                                        Mean.treated = mean(c(treat_1, treat_2, treat_3, treat_4), na.rm=T),
                                        SE.basal = std.error(c(basal_1, basal_2, basal_3, basal_4)),
                                        SE.treated = std.error(c(treat_1, treat_2, treat_3, treat_4)),
                                        Down.basal = Mean.basal - SE.basal,
                                        Up.basal = Mean.basal + SE.basal,
                                        Down.treated = Mean.treated - SE.treated,
                                        Up.treated = Mean.treated + SE.treated)
class(Data)

``` 

Gather values in long format
Change format to long and calculate mean+ and - SE in separate columns
```{r}
data.means <- Data[,c("Metabolites", "Mean.basal", "Mean.treated")] %>%
  gather(Sample, Mean, -Metabolites)
data.SE <- Data[,c("Metabolites",  "SE.basal", "SE.treated")] %>%
  gather(Sample, SE, -Metabolites)
data.UP <- Data[,c("Metabolites", "Up.basal", "Up.treated")] %>%
  gather(Sample, Up, -Metabolites)
data.DOWN <- Data[,c("Metabolites", "Down.basal", "Down.treated")] %>%
  gather(Sample, Down, -Metabolites)

dataMeanSe <- cbind(data.means,data.SE, data.UP, data.DOWN)
dataMeanSe$Sample <- factor(dataMeanSe$Sample)
dataMeanSe$Treatment <- factor(ifelse(grepl("basal", dataMeanSe$Sample), "basal","treated"))

#drop some unneeded columns
data.polar <- dataMeanSe[c(1,3,6,9,12, 13)]
```


To generate the polar plot, first we work with a normal line plot and add error bars as ribbon (or any style you want)
```{r}

a <- data.polar %>% 
  ggplot(aes(x=Metabolites, y = Mean, group=Treatment)) +
  #geom_boxplot(aes(group = Group, colour = Genotype, fill = Treatment, alpha = 0.2))
  geom_ribbon(aes(ymin = Down, ymax = Up, fill = Treatment), alpha = 0.15) +
  scale_fill_manual(values= c("blue", "red")) +
  geom_line(aes(colour = Treatment)) +
  scale_color_manual(values= c("blue", "red")) +
  theme_light() +
  theme(panel.grid.minor = element_blank())  

plot(a)
```

Now we change coordinate system
```{r}
a + coord_polar() 
```

